---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paleopop

<!-- badges: start -->
[![R-CMD-check](https://github.com/GlobalEcologyLab/paleopop/workflows/R-CMD-check/badge.svg)](https://github.com/GlobalEcologyLab/paleopop/actions)
<!-- badges: end -->

`paleopop` is an extension to `poems`, a spatially-explicit, process-explicit, pattern-oriented framework for modeling population dynamics. This extension adds functionality for modeling large populations at generational time-steps over paleontological time-scales. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("GlobalEcologyLab/paleopop")
```
## Example

One of the major additions in `paleopop` is the `PaleoRegion` R6 class, which allows for regions that change over time due to ice sheets, sea level, bathymetry, and so on. 

```{r paleoregion_example}
library(poems)
library(paleopop)
coordinates <- data.frame(x = rep(seq(-178.02, -178.06, -0.01), 5),
                          y = rep(seq(19.02, 19.06, 0.01), each = 5))
template_raster <- Region$new(coordinates = coordinates)$region_raster # full extent
sealevel_raster <- template_raster
template_raster[][c(7:9, 12:14, 17:19)] <- NA # make Ring Island
sealevel_raster[][c(7:9, 12:14, 17:18)] <- NA
raster_stack <- raster::stack(x = append(replicate(9, template_raster), sealevel_raster))
region <- PaleoRegion$new(template_raster = raster_stack)
raster::plot(region$temporal_mask_raster()[[1]], main = "Ring Island (first timestep)",
             xlab = "Longitude (degrees)", ylab = "Latitude (degrees)",
             colNA = "blue", legend = F)
raster::plot(region$temporal_mask_raster()[[10]], main = "Ring Island (last timestep)",
             xlab = "Longitude (degrees)", ylab = "Latitude (degrees)",
             colNA = "blue", legend = F)
```

`paleopop` also includes the `PaleoPopModel` class for running simulations over paleo time scales, and the `PaleoPopResults` class for storing the outputs from the paleo population simulator.

```{r paleosimulator_example}
model_template <- PaleoPopModel$new(
  region = region,
  time_steps = 10,
  years_per_step = 12, # years per generational time-step
  standard_deviation = 0.1,
  growth_rate_max = 0.6,
  harvest = F,
  populations = region$region_cells,
  initial_abundance = seq(9000, 0, -1000),
  transition_rate = 1.0,
  carrying_capacity = rep(1000, 17),
  dispersal = (!diag(nrow = 17, ncol = 17))*0.05,
  density_dependence = "logistic",
  dispersal_target_k = 10,
  occupancy_threshold = 1, 
  abundance_threshold = 10,
  results_selection = c("abundance")
)

results <- paleopop_simulator(model_template)
results # examine
raster::plot(region$raster_from_values(results$abundance[,10]),
             main = "Final abundance", xlab = "Longitude (degrees)", 
             ylab = "Latitude (degrees)", colNA = "blue")
```

A practical example of how to use `paleopop`, with more complex parameterization, can be found in the vignette.
